Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-11 write-ups/PLDI/paper.tex
Surgery

\ADComment{Revise: we're folding this into the overview of the examples.}

We remark on some issues that programmer working with cooperative
kernels should be aware of.

@@

@@

This guarantee can be exploited by algorithms that perform strided
data processing.  \ADComment{Tyler, can we give an example of a data
  processing loop that would have to be changed?} \TSComment{Done,
its in figure~\ref{fig:cgraphsearch}}

\paragraph{Placing the new constructs}

It is up to the programmmer to decide where to place our new language
constructs.  This must be done bearing in mind both the semantics of
the constructs, as detailed in Section~\ref{ }, as well as the
non-functional requirements of cooperative kernels that we describe in
Section~\ref{sec:nonfunctional}.

We note, however, that the programming model and associated language
constructs were inspired by the requirements of the irregular GPU
algorithms proposed in the literature, and that at least for these
algorithms it is usually easy to identify where to place the
constructs.

The irregular algorithms we encountered either employ work stealing or
operate on graph data structures.

Work stealing algorithms have a transactional flavour, whereby tasks
are repeatedly pulled from a queue and processed by a workgroup.  The
point of task completion---the end of a transaction---is a natural
place for a workgroup to invoke both $\offerkill$ and $\offerfork$, to
indicate both that the workgroup could gracefully exit computation,
but also that there is more work to be processed and that further
workgroups would be useful in order to process this work.  Such
algorithms usually require little state to persist between tasks;
typically only the pointers to the queue data structures need to be
pre-initialised, by annotating them with $\transmit$.
\ADComment{Refer back to cooperative versions of examples in
  Section~\ref{sec:semantics}.}

The level-by-level approach employed by many graph algorithms tends to
involve global barrier synchronisation.  In some cases, a global
barrier is required in the midst of a transaction, in which case
$\globalbarrier$, but not $\resizingglobalbarrier$, is suitable for
achieving synchronization without disturbing the set of active
workgroups.  But often it is the case that on completing a level of
the graph algorithm, the next level could be processed by more or
fewer workgroups, something which $\resizingglobalbarrier$
facilitates.  Again, only a small amount of state is relevant to new
workgroups joining the computation, such as the value of the level
being processed; this relevant state must be annotated with
$\transmit$.

\ADComment{Fold this in:}

Using coopeative kernels requires some programmer effort

\ADComment{Lifted:} , after which
we explain how the work stealing and graph traveral algorithms of
Section~\ref{sec:openclexamples} can be turned into cooperative kernels
that use the extensions.

@@

    level++;
    swap(in_nodes, out_nodes);
    reset(out_nodes);
(*@\label{line:cgraph:resizing}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
\caption{Cooperative kernel port for Figure~\ref{fig:graphsearch}}\label{fig:cgraphsearch}
\end{figure}

\ADComment{TODO.  We want to show the examples from Section~\ref{sec:openclexamples} ported to use cooperative kernels.  We want to make it clear that it is easy to place the primitives.}

In practice, we found that transforming traditional blocking GPU
kernels into cooperative kernels was straightforward. For example,
Figure~\ref{fig:cgraphsearch} shows the graph traversal example of
Figure~\ref{fig:graphsearch} transformed into a cooperative kernel.
On line~\ref{line:cgraph:resizing}, we change the original barrier
into a resizing barrier. Several variables are marked to be
transmitted in the case of workgroups joining at the resizing barrier
(lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
\ref{line:cgraph:transmit3}): {\tt level} must be restored so that
new workgroups know which frontier they are processing; {\tt
  in\_nodes} and {\tt out\_nodes} must be restored so that new
workgroups know which of the node arrays are input and output. Lastly,
the static work distribution of the original kernel is no longer valid
in a cooperative kernel. This is because the stride (which is based on
$M$) may change after each resizing barrier
call. To fix this, we re-distribute the work after each resizing
barrier call by recomputing the thread id and stride
(lines~\ref{line:cgraph:rechunking1} and
\ref{line:cgraph:rechunking2}).

\TSComment{Can Hugues talk about work stealing here?}

\caption{Cooperative version of work stealing: the number of workpools
must be explicitely passed as arguments since the number of workgroups
is not a constant anymore}\label{fig:workstealing-cooperative}
\end{figure}

>>>>

Working with cooperative kernels requires some manual effort on the
part of the programmer: they must place the $\transmit$ qualifier and
use the new language primitives described in
\mysec\ref{sec:semantics}, and account for slight changes in the OpenCL semantics induced by these new features.

The programming model and associated language constructs were inspired
by the requirements of the irregular GPU algorithms proposed in the
literature, and that at least for these algorithms it is usually easy
to identify where to place the constructs.

We discuss the main semantic change induced by our programming
model---that the number of workgroups executing a kernel varies
dynamically---and illustrate our programming model and the impact of
this semantic change by adapting the work stealing and graph traversal
algorithms of \mysec\ref{sec:openclexamples} to use cooperative
kernels.

@@

\paragraph{Adapting work stealing}

\myfiglong\ref{fig:workstealing-cooperative} shows a cooperative
version of the work stealing example of \myfig\ref{...} that we
discussed in \mysec\ref{sec:openclexamples}.
\ADComment{TODO: go through with Hugues; where are the offer functions?}

\caption{Cooperative version of the work stealing example from \myfig\ref{fig:XXX}; the number of workpools
must be explicitly passed as arguments since the number of workgroups
is not a constant anymore}\label{fig:workstealing-cooperative}
\end{figure}

@@

This guarantee can be exploited by algorithms that perform strided
data processing.  We now illustrate this using the graph traversal motivating example.

\paragraph{Adapting graph traversal} \ADComment{Tyler, the original had two global barriers; the port has just one resizing barrier.  Is this right?}
\myfiglong\ref{fig:cgraphsearch} shows a cooperative version of the graph traversal kernel
of \myfig\ref{fig:graphsearch}, introduced in
\mysec\ref{sec:openclexamples}.  
On line~\ref{line:cgraph:resizing}, we change the original global barrier
into a resizing barrier. Several variables are marked to be
transmitted in the case of workgroups joining at the resizing barrier
(lines~\ref{line:cgraph:transmit1}, \ref{line:cgraph:transmit2} and
\ref{line:cgraph:transmit3}): $\keyword{level}$ must be restored so that
new workgroups know which frontier they are processing; $\keyword{in\_nodes}$ and $\keyword{out\_nodes}$ must be restored so that new
workgroups know which of the node arrays are input and output. Lastly,
the static work distribution of the original kernel is no longer valid
in a cooperative kernel. This is because the stride (which is based on
$M$) may change after each resizing barrier
call. To fix this, we re-distribute the work after each resizing
barrier call by recomputing the thread id and stride
(lines~\ref{line:cgraph:rechunking1} and
\ref{line:cgraph:rechunking2}).  This example exploits the fact that the cooperative kernel does not issue $\offerkill$ nor $\offerfork$ directly: \ADComment{Actually I'm confused and need to discuss :-)}

@@

    level++;
    swap(in_nodes, out_nodes);
    reset(out_nodes);
(*@\label{line:cgraph:resizing}@*)    resizing_global_barrier();
  }
}
\end{lstlisting}
\caption{Cooperative version of the graph traversal kernel of \myfig\ref{fig:graphsearch}, using a resizing global barrier and $\transmit$ annotations}\label{fig:cgraphsearch}
\end{figure}

In \mysec\ref{ } we describe the set of irregular GPU algorithms used
in our experiments, which largely captures the irregular blocking
algorithms that are available as open source GPU kernels.  These all
employ either work stealing or operate on graph data structures.
Like the simplified example of \myfig\ref{ }, the work stealing algorithms have a transactional flavour
and require little or no state to be carried between transactions.  Thus the point at which a workgroup is ready to process a new task is a natural point to place $\offerkill$ and $\offerfork$, and few or no $\transmit$ annotations are required.
%
The example of \myfig\ref{fig:cgraphsearch} is simpler than, but representative of,
most level-by-level graph algorithms.  These algorithms tend to
require more state to be annotated with $\transmit$, and in some cases
a global barrier is required in the midst of a transaction, in which
case $\globalbarrier$, but not $\resizingglobalbarrier$, is suitable
for achieving synchronization without disturbing the set of active
workgroups.  But often it is the case that on completing a level of
the graph algorithm, the next level could be processed by more or
fewer workgroups, something which $\resizingglobalbarrier$
facilitates.

In summary, our experience so far indicates that cooperative kernels
is a good fit for current irregular algorithms.  Of cousre this is, in
part, because we designed the model with these algorithms in mind, but
we argue that the fit is natural not just for the examples that we
study, but more generally for at least these two patterns of irregular
algorithm.
